---
title             : "Combined and relative contributions of risk and protective factors to mood and mood disorder in daily life"
shorttitle        : "Relative contributions to mood and mood disorder"
author: 
    
  - name          : "Vera E. Heininga"
    affiliation   : "1"
    corresponding : yes
    address       : "Grote Kruisstraat 2/1, 9712 TS  Groningen, The Netherlands"
    email         : "v.e.heininga@rug.nl."
  - name          : "Nienke Jonker"
    affiliation   : "2"
    corresponding : no
    address       : "Tiensestraat 102 - bus 3713, 3000 Leuven"
    email         : "n.jonker@rug.nl"
  - name          : "Bertus Jeronimus"
    affiliation   : "1"
    corresponding : no
    address       : "Tiensestraat 102 - bus 3713, 3000 Leuven"
    email         : "b.f.jeronimus@rug.nl"
  - name          : "Peter de Jonge"
    affiliation   : "1"  
    corresponding : no  
    address       : "Tiensestraat 102 - bus 3713, 3000 Leuven"
    email         : "peter.d.jonge@rug.nl"
affiliation:
    
  - id            : "1"
    institution   : "University of Groningen, Department of Developmental Psychology, The Netherlands"
  - id            : "2"
    institution   : "Department of Clinical Psychology and Experimental Psychopathology, University of Groningen, The Netherlands"
abstract: |
  
  
  **Background**: Sleep quality, physical activity, social interaction, negative events, and coping are the five most robust risk and protective factors for mood and mood disorders in daily life. Despite that these factors are all part of everyday life, the impact of these factors have been almost exclusively investigated in isolation, leaving their combined and relative impact on mood and mood disorders largely unknown. The present study used ecological momentary assessment (EMA) methods to: (1) replicate the combined and relative contribution of these five factors to negative affect (NA); (2) do a first examination of the collective and relative contribution of these five to Positive Affect (PA); and (3) explore differences in the combined and relative contribution to NA and PA in people with low and high levels depressive symptoms. 
keywords          : "Depression; Anhedonia; Positive Affect; individual differences; Ecological Momentary Assessment (EMA)"
wordcount         : 
bibliography      : ["references.bib"]
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
figsintext        : no
documentclass     : "apa6"
mask              : no
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  chunk_output_type: inline\
---

```{r ReadMe, include=FALSE}

# Preregistration doi: https://osf.io/8n54h
# Preprint doi:

# Below you will find the programming work with regard to:

# Preperations:
    # Preparing and merging relevant data

# Variable computations for hypotheses testing:
    # Variables beloning to H1:  a positive association between stressful events experienced in the past six hours and subsequent changes in NA, 
    #   and a negative association between coping ability in the past six hours and subsequent changes in NA. [directional]
    # Variables beloning to H2: a positive association between appraisal of social company in the past six hours and subsequent changes in PA, 
    #    and a positive association between physical activity in the past six hours and subsequent changes in PA. [directional] 

# Please, download Data and Materials from OSF: https://osf.io/drczs/
# If you cannot reproduce our variables and/or findings in R, or if you have questions, 
# critics, or compliments, you could try v.e.heininga@rug.nl, or v.e.heininga@gmail.com!

```

```{r load_packages, include=FALSE}

if (!require("pacman")) install.packages("pacman", repos ="http://cran.us.r-project.org")
pacman::p_load(foreign, summarytools, dplyr, expss, remotes)


# disable scientific notation in the output to make it easy to read
options(scipen=999)

```

```{r data, message=FALSE}
Data <- read.spss('x:\\2019\\Papers\\Relative contributions\\3_Data\\200604_HGIN097_Heininga_Diary181219.sav', to.data.frame=TRUE, use.value.labels=FALSE)

library(dplyr)
MyData <- Data %>% select(patient_id, sex, age_diary, QIDSsum_prediary,
                          # Diary items for descriptives and analyses:
                          Real_period, meting, Observations_valid90, time, morning, afternoon, evening, time_completed, mad_diary_33,
                          # NA items:
                          mad_diary_6, mad_diary_8, mad_diary_10, mad_diary_12, mad_diary_14, mad_diary_16,
                          # PA items:
                          mad_diary_5, mad_diary_7, mad_diary_9, mad_diary_11, mad_diary_13, mad_diary_15,
                          # Risk/protective factors:
                          mad_diary_3, mad_diary_25, mad_diary_32, mad_diary_37, mad_diary_41
                          )

# labeling in accordance with Table2 from Krieken et al (2016):

library(expss)
MyData = apply_labels(MyData,
                      patient_id = "participant identifier for the diary study", 
                      sex = "binary version of gender variable from the cross-sectional part of HoeGekIsNL", 
                      age_diary = "age in years, based on birth year from the cross-sectional part of HoeGekIsNL. Time-varying variable.", 
                      QIDSsum_prediary = "sum scores from the pre-diary measurement on the QIDS",
                      Real_period = "the period with most valid observations for each meting", 
                      Real_period = c("assessment belongs to period with most valid observations" = 1,
                                      "assessment belongs to not a period with most valid observations" = NA), 
                      meting = "Number of 30 day diary study the particpant participated in. That is, if participants received feedback after the 30 days of diary study, they could start a new 30 day diary study.", 
                      Observations_valid90 = "total number of valid observations (within 1 hour) (max 90)",
                      time = "time point, or assessment number 1-90", 
                      morning = c("No" = 0,
                                  "Yes"=1), 
                      afternoon = c("No" = 0,
                                  "Yes"=1), 
                      evening = c("No" = 0,
                                  "Yes"=1), 
                      time_completed = "completion time in minutes", 
                      mad_diary_33 = "Context of special event. This was related to:",
                      mad_diary_33 = c("Myself" = 1,
                                      "Home situation/ close family/ significant others" = 2,
                                      "Friends/ other family/acquaintances" = 3,
                                      "Work/ school" = 4,
                                      "Society/ news" = 5,
                                      "Public space/ strangers" = 6,
                                      "Other" = 7),
                      mad_diary_6 = "I feel gloomy", 
                      mad_diary_8 = "I feel anxious", 
                      mad_diary_10 = "I feel nervous", 
                      mad_diary_12 = "I feel irritable", 
                      mad_diary_14 = "I feel dull", 
                      mad_diary_16 = "I feel tired",
                      mad_diary_5 = "I feel relaxed", 
                      mad_diary_7 = "I feel energetic", 
                      mad_diary_9 = "I feel enthusiastic", 
                      mad_diary_11 = "I feel content", 
                      mad_diary_13 = "I feel calm", 
                      mad_diary_15 = "I feel cheerful",
                      mad_diary_3 = "Did you sleep well? “Not at all” to “Very well” (0-100)", 
                      mad_diary_25 = "I am unbalanced, “Not at all” to “Very much” (0–100)", 
                      mad_diary_32 = "Did something special happen since the last measurement?", 
                      mad_diary_32 = c("No, nothing" = 1,
                                      "Yes, something positive" = 2,
                                      "Yes, something neutral" = 3,
                                      "Yes, something negative" = 4),
                      mad_diary_37 = "Appraisal of social context, “Very unpleasant”, via “Neutral”, to “Very pleasant” (0-50-100)", 
                      mad_diary_41 = "Physical activity, “Not at all” to “Very much” (0–100)"
)


# The ReadMe file of the HGNL study instructs data analysts to "Be sure to select Real_period= 1 for your analyses!" and to "Filter meting=1 & Real_period=1 & time<=90 should be sufficient when you do analyses on the diary items for the first official diary study (meting=1)."

MyData_filter <- MyData %>% filter(meting == 1 & Real_period == 1 & time < 90 )

## Data preparations for paper

# PA and NA
MyData_filter <- MyData_filter %>% mutate(
                                        PosAffect = (mad_diary_5 + mad_diary_7 + mad_diary_9 + mad_diary_11 + mad_diary_13 + mad_diary_15) / 6,
                                        NegAffect = (mad_diary_6 + mad_diary_8 + mad_diary_10 + mad_diary_12 + mad_diary_14 + mad_diary_16) / 6
                                        )

# Negative events:  with answer categories being (1) No, nothing; (2) Yes, something positive; (3) Yes, something neutral; (4) Yes, something negative. (please note: will be binary recoded to reflect negative events, see section "Transformations" for details)

# Creating descriptive variables
remotes::install_github("wviechtb/esmpack")
library(esmpack)

# Calculate within subject mean PA
MyData_filter$PosAffect_m <- calc.mean(PosAffect, patient_id, data=MyData_filter, expand=TRUE)

# Calculate within subject mean NA
MyData_filter$NegAffect_m <- calc.mean(NegAffect, patient_id, data=MyData_filter, expand=TRUE)

# Calculate mean-centered values of relevant variables for each subject
# note: MyData_filter$x_c <- calc.mcent(x, id, data, na.rm=TRUE)
# see: https://rdrr.io/github/wviechtb/esmpack/man/calc.mean.html
# and see: http://stulp.gmw.rug.nl/OSCG/transformdata.html

# note for later to note in methods:
# Following the van Roekel guidelines
# : 18:00 was the earliest possibility for the evening measurement, 6:00 was the latest possibility).
```

# Background
Risk and protective lifestyle factors have been almost exclusively investigated in isolation, leaving their *combined and relative impact* on mood and mental health largely unknown. Identifying such impact can inform prevention and treatment strategies, and can help to uncover underlying mechanisms. 

Building upon the risk and protective factors from daily life identified by Pemberton & Fuller-Tyszkiewicz (2016), Fuller-Tyszkiewicz and colleagues (2017) investigated the collective and relative contributions of the five most robust unidirectional lifestyle factors (i.e., sleep quality, appraisal of social company, negative event, coping, and physical activity) to subsequent changes in **Negative Affect**. 

In the present study, using a larger sample and a fixed Experience Sampling Method (ESM) study design of 3 time points a day for 30 consecutive days, we replicate part of the analyses of Fuller-Tyszkiewicz and colleagues (2017) with regard to Negative Affect, and extend their finding with a first description of the within-person combined and relative contributions of the five risk/protective factors for **Positive Affect** in daily life.

## The present study
We expect that a single regression model that includes sleep quality, appraisal of social company, negative event, coping, and physical activity, will show:

1. Similar outcomes as Fuller-Tyszkiewicz and colleagues (2017) reported, namely a positive association between stressful events and Negative Affect several hours later, and a negative association between coping abilities and Negative Affect approximately several hours later, with perceived stress as the strongest predictor of change in **Negative Affect**, followed by prior level of negative mood and poor coping ability. [directional]

2. A positive association between appraisal of social company and **Positive Affect** approximately several hours later, and a positive association between physical activity and **Positive Affect** approximately several hours later. [directional] 

# Methods

## Sample
The sample used is a subsample of the diary part of the HowNutsAreTheDutch (HND; Dutch: HoeGekIsNL; van der Krieke et al., 2015; 2016). HND is a national crowdsourcing study designed to investigate multiple continuous mental health dimensions in a sample from the general population (n > 16.000). Participants can enroll themselves on an internet platform that allows participants to (a) compare themselves to other participants via cross-sectional questionnaires and (b) to monitor themselves three times a day for 30 days with an intensive longitudinal diary study via their smartphone.

HND launched an open call to inhabitants of the Netherlands to join the research in December of 2013, and invited them to visit the Dutch website www.HoeGekIs.nl (also www. HowNutsAreTheDutch.com). The open call was announced on local and national radio broadcasts, television, during local podium discussions, in newspapers, and in magazines. The news about the HND research project was further disseminated via personal outlets of those interested (e.g., online blogs, twitter). To participate, people had to register, create, and confirm their account on www.HoeGekIs.nl. In a general start questionnaire, participants were asked to provide information such as gender, birth year, and country of residence (the Netherlands/Belgium/Other). Although HND was initially targeted on Dutch citizens located in The Netherlands, a question on country of residence was included after news about the study spread to Belgium and other Dutch speaking regions.

### Inclusion rules
The study requirements for the larger HND study were age 18 or above; having a mobile phone with internet connection; not engaged in shift work; and not anticipating a major disruption of daily routines (e.g. a planned trip abroad, an anticipated surgical operation). In addition, participants were also asked to indicate whether they were aware that participation would be useless in case too many assessments would be missed, and whether they approved that their anonymized data would be used for scientific research. Participants had to check a box for each of these requirements before they could proceed.

### Procedure diary study
After enrollment, before the diary part started, participants had to complete a baseline assessment consisting of the items of the positive and negative affect schedule (PANAS; Peeters et al., 1996; Raes et al., 2009), the Quick Inventory of Depressive Symptoms (QIDS; Rush et al., 2003, 2006), and two extra items retrieved from the Inventory of Depressive Symptomatology (IDS; Rush et al., 1996) to assess anxiety/panic symptoms. 

From the start of the diary study, for 30 days, participants filled out a short questionnaire of 43 diary questions on a variety of topics (e.g., subjective well-being, sleep, mood, mindfulness, and being outdoors) on their smartphone three times a day with a fixed 6 hour interval Experience Sampling Method (ESM) design. To lower the burden of the diary study, the starting time of the assessment schedule was chosen in colsultation with the participant. For example, people getting up at 8:00 could request their first diary at 8:30. Participants received a text message on their mobile phone with a link to a questionnaire, and were asked to fill out the questionnaire as soon as possible within one hour. After one hour, the questionnaire was no longer be accessible.

## Measures
### Negative Affect
Based on the circumplex model of affect, which describes valence and activation dimensions of affect (Barrett and Russell, 1998; Yik et al., 1999), the NA variable reflect the mean of the following momentary affect-items that participants rated three times a day on a slider scale ranging from "Not at all" (0) to "Very much" (100):

1. I feel gloomy
2. I feel anxious
3. I feel nervous
4. I feel irritable
5. I feel dull
6. I feel tired

### Positive Affect
In a similar vein, the PA variable reflect the mean of the following momentary affect-items that participants rated three times a day on a slider scale ranging from "Not at all" (0) to "Very much" (100):

1. I feel relaxed
2. I feel energetic
3. I feel enthusiastic
4. I feel content
5. I feel calm
6. I feel cheerful

### Sleep quality 
Participants were asked "Did you sleep well?", with an answer-slider ranging from “Not at all” to “Very well” (0-100).

### Appraisal of social company
Participants were asked "I found my company predominantly", with an answer-slider ranging from “Very unpleasant”, via “Neutral”, to “Very pleasant” (0-50-100).

### Negative event
Participants were asked "Did something special happen since the last measurement?" with answer categories being (1) No, nothing; (2) Yes, something positive; (3) Yes, something neutral; (4) Yes, something negative. The ansers were recoded reflecting 0 when nothing special happened since the last measurement or something positive or something neutral (i.e., 0 = 1/2/3), and 1 when a negative event took place (i.e., 1 = 4).

### Coping
Participants were asked "I am unbalanced", with an answer-slider ranging from “Not at all” to “Very much” (0–100).

### Physical activity
Since the last measurement I was physically active, answered on a slider ranging from “Not at all” to “Very much” (0–100)

## Statistical procedures
We investigated our hypotheses in R `r getRversion()` [Version 3.5.1; @R], and use R Markdown [@Rmarkdown] and papaja [@papaja] to create a reproducible manuscript. All hypotheses were preregistered on Open Science Framework (insert link).  

To test our hypotheses, two separate random slopes Linear Mixed-Effects Models were used for respectively Negative Affect and Positive Affect with assessments nested within subjects. All predictors were person-mean centered, except those of which the zero-point is already meaningful on its own and present in the data (e.g., negative events). Both concurrent associations (measured at the same time point; Equation 1 and Equation 3-8), and lagged associations were examined (measured six hours apart; Equation 2-8):

Concurrent associations, Level 1 (Equation 1):
$$Affect_{i}=\beta_{0}+\beta_{1}(Sleep_{i})+\beta_{2}(Event_{i})+\beta_{3}(Coping_{i})+\beta_{4}(Physical_{i})+\beta_{5}(Social_{i})$$

Lagged associations, Level 1 (Equation 2):
$$Change \ in \ affect_{(i-(i-1))}=\beta_{0}+\beta_{1}(Sleep_{i-1})+\beta_{2}(Event_{i-1})+\beta_{3}(Coping_{i-1})+\beta_{4}(Physical_{i-1})+\beta_{5}(Social_{i-1})$$

Lagged associations, Level 2 (Equation 3-8):
$$
\begin{aligned}
  \beta_{0}&=\gamma_{00}+\gamma_{01}+u_{0}			\\
  \beta_{1}&=\gamma_{10}+\gamma_{11}+u_{1}      \\
  \beta_{2}&=\gamma_{20}+\gamma_{21}+u_{2}      \\
  \beta_{1}&=\gamma_{30}+\gamma_{31}+u_{3}      \\
  \beta_{2}&=\gamma_{40}+\gamma_{41}+u_{4}      \\
  \beta_{1}&=\gamma_{50}+\gamma_{51}+u_{5}      
\end{aligned}
$$

# Results
## Descriptive statistics


### Distributions of main variables

```{r means, message=FALSE, warning=FALSE}

meanAnhedonia<- round(mean(MyData_Table2$Anhedonia, na.rm=TRUE),2)
sdAnhedonia<- round(sd(MyData_Table2$Anhedonia, na.rm=TRUE),2)

meanQuality<- round(mean(MyData_Table2$Poor_relationship_quality, na.rm=TRUE),2)
sdQuality<- round(sd(MyData_Table2$Poor_relationship_quality, na.rm=TRUE), 2)

meanAbandonment<- round(mean(MyData_Table2$Fear_of_abandonment , na.rm=TRUE),2)
sdAbandonment<- round(sd(MyData_Table2$Fear_of_abandonment , na.rm=TRUE),2)
```

The average mean level of within-person negative affect was `r mean` (between-person *SD*=`r sd`); pos affect + five predictors.

### Intraclass Correlation Coefficient (ICC) of affect

```{r descriptives_ICC, message=FALSE, warning=FALSE}
# To describe how much variation in momentary Anhedonia comes from differences between versus within individuals, we calculate the Intraclass correlation coefficient (ICC) by the following empty or null model of momentary Anhedonia  (Anhedonia ):

library(lme4)
library(lmerTest)
library(sjstats)

ICC_model <- lmerTest::lmer(Anhedonia ~ 1 + 
                (1 | ID_anonym), 
        data =MyData)

ICC <- icc(ICC_model)

ICC.prop <- round(ICC$ICC_adjusted*100, digits=0) 

ICC.prop2 <- (100-ICC.prop)

# See for details: http://glmm.wikidot.com/faq
dat.new <- data.frame(Time=1:10)
dat.new$Anhedonia <- predict(ICC_model, dat.new, re.form=NA)
m.mat <- model.matrix(terms(ICC_model), dat.new)
dat.new$var <- diag(m.mat %*% vcov(ICC_model) %*% t(m.mat)) + VarCorr(ICC_model)$ID_anonym[1]
dat.new$pvar <- dat.new$var + sigma(ICC_model)^2
dat.new$ci.lb <- with(dat.new, Anhedonia - 1.96*sqrt(var))
dat.new$ci.ub <- with(dat.new, Anhedonia + 1.96*sqrt(var))
dat.new$pi.lb <- with(dat.new, Anhedonia - 1.96*sqrt(pvar))
dat.new$pi.ub <- with(dat.new, Anhedonia + 1.96*sqrt(pvar))

```

The ICC was `r round(ICC$ICC_adjusted, digits=3)`, indicating that `r ICC.prop2`% of the observed variance in affect stems from within-person differences.

### Diurnal pattern of affect

```{r descriptives_diurnal, message=FALSE, warning=FALSE}

# To describe the course of variation in Anhedonia levels, we will estimate a three-level hierarchical structure, we will investigate the linear and quadratic effects of the number of minutes that the assessment took place after 9:30 (*t*), nested within days (day), and nested within subjects (ID_anonym) using random intercepts for subjects and days nested within subjects:

diurnal <- lmerTest::lmer(Affect ~ Time + I(Time^2) + 
                 (Time + I(Time^2) | ID_anonym:day) + 
                 (Time + I(Time^2) | ID_anonym),
        data =MyData)

diurnal.sum <- summary(diurnal)

# extract coefficients linear
diurnal.est <- printnum(diurnal.sum$coefficients[2], digits=2) # gives difference in intercept by group
diurnal.est.t <- round(diurnal.sum$coefficients[11], digits=2) # t-value
diurnal.est.p <- printp(diurnal.sum$coefficients[14]) # one-tailed p-value

# extract coefficients Quadratic
diurnal.est2 <- round(diurnal.sum$coefficients[3], digits=2) # gives difference in intercept by group
diurnal.est2.t <- round(diurnal.sum$coefficients[12], digits=2) # t-value
diurnal.est2.p <- printp(diurnal.sum$coefficients[15]/2) # one-tailed p-valu

```

Because affect often follow a diurnal pattern, we explored linear trends and quadratic trends. There was no linear trends of affect within days (*B*=`r diurnal.est`; *t* =`r diurnal.est.t`; *p* =`r diurnal.est.p`), nor a quadratic trend (*B*=`r diurnal.est2`; *t* =`r diurnal.est2.t`; *p* =`r diurnal.est2.p`).


### Correlation between risk/protective factors

```{r correlation, message=FALSE, warning=FALSE}
# Exploratory: Correlation betweenthe two social stress items 
library(rmcorr)
my.rmc <- rmcorr(participant = ID_anonym, measure1 = Fear_of_abandonment_c, measure2 = Poor_relationship_quality_c, dataset = MyData)
plot(my.rmc, overall = TRUE)

```


## H1: Combined and relative contributions to Negative Affect 

```{r assumptionsH1, message=FALSE, warning=FALSE}
assumptions

```

## H2: Combined and relative contributions to Positive Affect 


## Combined and relative contributions in depression

### Exploratory posthoc analyses
To meet the criteria for depression, in additon to the total score of 6, particpants should at least experience mild anhedonia (i.e., scored “I notice that I am less interested in people or activities." or worse) and/or mild sad mood (i.e., scored “I feel sad less than half the time." or worse).

## Validity checks 
repeat our main analyses among the subsample of people who completed > 70% of the ESM assessments

## Deviations from pre-registration (link)
1. had to use filter meting=1 & Real_period=1 & time<=90. 

## Additional exploratory posthoc analyses

1. all in one model with PA and NA as outcome
2. *Changes* in x are associated with changes in affect thereafter?
3. bidirectional lagged effects?
4. relative standard deviation to avoid floor or ceiling effects (Mestdagh et al., 2018): Mestdagh, M., Pe, M., Pestman, W., Verdonck, S., Kuppens, P., & Tuerlinckx, F. (2018). Sidelining the mean: The relative variability index as a generic mean-corrected variability measure for bounded variables. Psychological Methods, 23(4), 690–707. https://doi.org/10.1037/met0000153
5. lagged associations

# Discussion


## Study limitations
This study has several limitations. First, (...)

# Conclusions



# Declarations

## Author's contribution
V.E.H. and N.J. developed the study concept. Data collection was performed by B.F.J and P.d.J., and V.E.H. performed the data analysis. V.E.H. drafted earlier versions of the manuscript, and N.J., B.F.J. and P.d.J. provided critical comments and suggestions for revisions. All authors approved the final version of the paper for submission.

## Ethics approval and consent to participate


## Funding



\newpage

# References

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

